---
title: "Rare Variant Preparation and FGSEA"
author: "Amalya Murrill"
date: "`r Sys.Date()`"
output: github_document
 
---
This script performs fgsea pathway enrichment analysis on both the rare variant (Satterstrom et al. 2020 TADA+ q-values) and common variant (MAGMA Z-statistics) arms. A harmonization problem exists between the two summary statistics; as a provisional measure, rare variant genes below the first quartile of -log10(q-values) were retained to reduce zero-inflation. Gene IDs are standardized to Entrez, ranked, and fed into fgsea using ASD-relevant GO-BP and KEGG pathway sets. Convergent and divergent pathways are identified through qualitative overlap and a MAD-based quantitative classification on NES differences. An FDR threshold of 0.25 serves as a placeholder pending formal harmonization. Classified pathways are exported for downstream gene-level and constraint analysis.        

```{r knitting-rules}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(org.Hs.eg.db)
library(fgsea)
library(msigdbr)
library(patchwork)
```

## Data Import: Rare Variants

```{r rare-read-in}
rare <- read_tsv("../data/raw/ASC_gene_results.tsv.bgz", col_names= TRUE)
#quick check
head(rare) # two junk gene names
colnames(rare)
nrow(rare) # 17247 genes

#remove junk gene names
rare <- rare %>% filter(startsWith(gene_id, "ENSG"))
head(rare)
nrow(rare) # 17345, 2 bad genes removed
```

## Data Import: Common Variant MAGMA Output

```{r common-magma-read-in}
magma_common <- read_table("../data/processed/common_MAGMA_genes.genes.out")
#quick check
head(magma_common)
colnames(magma_common)
nrow(magma_common) # 18072 genes
```

## Visualize Distributions

```{r stat-distributions}
#plot the rare
rare_dist <- rare %>% ggplot(aes(x = qval)) +
  geom_histogram(fill = "#F4D03F",alpha = 0.9, bins = 30) +
  xlab("TADA+ q-value") +
  theme_classic()

rare_dist

#plot the common
common_dist <- magma_common %>% ggplot(aes(x = ZSTAT)) +
  geom_histogram(fill = "lightblue", alpha = 0.9, bins = 50) +
  xlab("MAGMA Z Statistic") +
  theme_classic()

common_dist

#put them together
distributions <- common_dist + rare_dist

distributions


#plot -log10rare
rare_log_quants <- rare %>% dplyr::select(qval) %>% pull() %>% quantile()

rare_dist_log <- rare %>% ggplot(aes(x = -log10(qval))) +
  geom_histogram(fill = "#F4D03F",alpha = 0.9, bins = 80) +
  geom_vline(xintercept = rare_log_quants, color = "grey", linetype = "dashed" ) +
  xlim(0.5, 2.5) +
  xlab("TADA+ -log10(q-value)") +
  theme_classic()

rare_dist_log # remove everything before first quartile
rare_log_quants
#-log10(q) = 0.87522, so 
qval_thresh <- 10^(-0.87522) # 0.13
```

### Chi-Squared Goodness of Fit

```{r rare-chi-square}

#use chi-square GOF to see threshold of deviation from uniform distribution in this range 
#qval_start <- rare %>% arrange(qval) %>% dplyr::select(qval) %>% pull()
#test <- list(p.value = 0)
#threshold <- 0.05

#while(test$p.value < 0.05 & threshold < 0.5){
#threshold <- threshold + 0.01
#qval_current <- qval_start
#keep <- qval_current <= threshold
#qval_current <- qval_current[keep]
#binned_qvals<- cut(qval_current, breaks = 50)
#observed <- table(binned_qvals) 
#test <- chisq.test(observed)
#if(test$p.value > 0.05){
  #print(paste0("threshold =",threshold))
#}
#}

```

## Apply Data-Driven Filter to Rare Arm

```{r filter-by-quantile}
rare_filt <- rare %>% filter(qval < qval_thresh)
#quick check
nrow(rare_filt) # from 17247 genes to 118
head(rare_filt)
```

## Make Rare Ranked Vector

```{r rare-filter-and-rank}
rare_tib <- rare %>% mutate(neglog10q = -log10(qval)) %>% dplyr::select(gene_id, neglog10q) %>% arrange(desc(neglog10q))
rare_tib 
# look at data to check for NAs or Inf before making ranked vector
sum(is.na(rare_tib)) # 0 NA values
sum(is.infinite(rare_tib$neglog10q)) # 3 Inf values
#go back and correct for Inf values using qval floor of 1e-300 (winsorize)
rare_tib_adj <- rare %>% mutate(neglog10q = -log10(pmax(qval, 1e-300))) %>% dplyr::select(gene_id, neglog10q) %>% arrange(desc(neglog10q))

head(rare_tib_adj)

#now rank for fgsea
rare_ranked <- deframe(rare_tib_adj)
head(rare_ranked)

#convert to Entrez IDs for fgsea
entrez_ids <- mapIds(
  org.Hs.eg.db,
  keys = names(rare_ranked),
  column = "ENTREZID",
  keytype = "ENSEMBL",
  multiVals = "first"
)

#how many are NA?
sum(is.na(entrez_ids)) # 110

#subset by non-NA values
keep_idx <- !is.na(entrez_ids)
entrez_ids <- entrez_ids[keep_idx]
rare_ranked <- rare_ranked[keep_idx]
#rename
names(rare_ranked) <- entrez_ids

#deduplicate
rare_ranked_df <- enframe(rare_ranked, name = "gene_id", value = "neglog10q") %>% group_by(gene_id) %>% slice_max(neglog10q, n = 1, with_ties = FALSE) %>% ungroup()

rare_ranked <-deframe(rare_ranked_df)
  
rare_ranked
```

## Make Common Ranked Vector

```{r common-filter-and-rank}
common_tib <- magma_common %>% dplyr::select(GENE, ZSTAT) %>% arrange(desc(ZSTAT)) #use ZSTAT to preserve direction
head(common_tib)

# look at data to check for NAs or Inf before making ranked vector
sum(is.na(common_tib)) # 0 NA values
sum(ZSTAT = 0) # no 0 ZSTAT values

#now rank for fgsea
common_ranked <- deframe(common_tib)
head(common_ranked)
```


## Pathway Selection

```{r select-pathways}
go_bp <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")
kegg <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "KEGG")

head(go_bp)
head(kegg)

# filter pathways by canonical terms: SYNAP, GABA, EXCIT, INHIB, CHROMATIN, NEURO, AXON, GLUTAMAT, MTOR (metabolism)

name_filt_pattern <- "SYNAP|GABA|EXCIT|INHIB|CHROMATIN|NEURO|AXON|GLUTAMAT|MTOR"
go_bp_filt <- go_bp %>% filter(grepl(name_filt_pattern, gs_name))
head(go_bp_filt)

kegg_filt <- kegg %>% filter(grepl(name_filt_pattern, gs_name))
head (kegg_filt)
```

## List Format Conversion

```{r msigdbr-lists}
# merge two dataframes and make list
pathways <- rbind(go_bp_filt, kegg_filt) %>% split(x = as.character(.$entrez_gene), f = .$gs_name)
head(pathways)
```


## FGSEA

```{r rare-fgsea}
rare_fgseaRes <- fgsea(
  pathways = pathways,
  stats = rare_ranked,
  minSize = 15,
  maxSize = 500,
  scoreType = c("pos") # from TADA+ qvals, all positive
)

nrow(rare_fgseaRes[order(pval), ]) # 182 

rare_pathways <- rare_fgseaRes[order(pval), ]
```


### Collapse Rare Variant Pathways

```{r rare-collapse}
rare_collapsed <- fgsea::collapsePathways(
  fgseaRes = rare_fgseaRes,
  pathways = pathways,
  stats = rare_ranked
)
rare_collapsed$mainPathways

write.csv(as.data.frame(rare_collapsed$parentPathways), "../results/rare_collapsed_pathway_mapping.csv") #for summary
```


```{r common-fgsea}
common_fgseaRes <- fgsea(
  pathways = pathways,
  stats = common_ranked,
  minSize = 15,
  maxSize = 500
)

nrow(common_fgseaRes[order(pval), ]) # 184

common_pathways <- common_fgseaRes[order(pval), ]
```

### Collapse Common Variant Pathways

```{r common-collapse}
common_collapsed <- fgsea::collapsePathways(
  fgseaRes = common_fgseaRes,
  pathways = pathways,
  stats = common_ranked
)
common_collapsed$mainPathways

write.csv(as.data.frame(common_collapsed$parentPathways), "../results/common_collapsed_pathway_mapping.csv") # for summary
```

### Visualize Pathway Statistics

```{r plot-pathway-padjs}

#plot the rare paths
rare_path_padjs <- rare_pathways %>% ggplot(aes(x = padj)) +
  geom_histogram(fill = "lightpink",alpha = 0.9, bins = 30) +
  xlab("Rare Pathway Padj") +
  theme_classic()

#plot the common paths
common_path_padjs <- common_pathways %>% ggplot(aes(x = padj)) +
  geom_histogram(fill = "lightgreen", alpha = 0.9, bins = 50) +
  xlab("Common Pathway Padj") +
  theme_classic()


#put them together
pathway_padjs <- common_path_padjs + rare_path_padjs

pathway_padjs
```


## Combine Pathway Findings

```{r combine-paths}
rare_pathways <- rare_pathways %>% mutate(variant = rep("rare", nrow(rare_pathways))) %>% relocate(variant)
common_pathways <- common_pathways %>% mutate(variant = rep("common", nrow(common_pathways))) %>% relocate(variant)
all_paths <- rbind(rare_pathways, common_pathways)

#are common and rare both there?
head(all_paths)
tail(all_paths)

#QC
nrow(all_paths) # 366
sum(is.na(all_paths)) # 0
```

## Qualitatively Overlap Pathways

```{r qualitative-overlap}
common_paths_main <- all_paths %>% filter(variant == "common") %>% pull(pathway)
rare_paths_main <- all_paths %>% filter(variant == "rare")  %>% pull(pathway)
overlap <- common_paths_main %in% rare_paths_main
overlap_paths <- common_paths_main[overlap]
overlap_paths # 182
not_in_overlap <- common_paths_main[!overlap]
not_in_overlap # 2 divergent by definition
```

```{r filter-by-overlap}
all_paths_overlap <- all_paths %>% filter(pathway %in% overlap_paths)
all_paths_overlap
nrow(all_paths_overlap) # 364
all_paths_overlap %>% filter(variant == "rare") %>% nrow() # 182
all_paths_overlap %>% filter(variant == "common") %>% nrow() # 182
```

## Visualize NES Distribution Calculate Difference in NES 

```{r visualize-NES}
NES_hist <- ggplot(all_paths_overlap, aes(x = NES, fill = variant)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  scale_fill_manual(values = c("lightblue", "#F4D03F"))+
  theme_classic()

NES_hist # shows right skew for rare, will use MAD for defining convergence

#ggsave("NES_distributions.png", plot = NES_hist, path = "../figures", width = 14, height = 10, dpi = 300)
```

```{r pivot-wider-NES-padj}
variant_path_NES_padj <- all_paths_overlap %>%
  dplyr::select(variant, pathway, NES, padj) %>%
  pivot_wider(
    names_from = variant,
    values_from = c(NES, padj),
    names_glue = "{variant}_{.value}"
  )

variant_path_NES_padj
```

## Calculate Unfiltered MAD

``` {r MAD-unfiltered-calc}
variant_path_NES_padj <- variant_path_NES_padj %>% mutate(
  NES_diff = rare_NES - common_NES, 
  MAD_unfilt = mad(NES_diff),
  median_NES_diff = median(NES_diff)) %>% relocate(NES_diff, .before = MAD_unfilt)
variant_path_NES_padj

#unfiltered MA for robustness check: 0.5844998
```

## Unfiltered Convergence (or Divergence) Classification

```{r classify-unfilt}
#threshold of 1 MAD away from the median for divergence:

#unfiltered median ± 1 MAD = 0.75 ± 0.58
#unfiltered convergence zone: 0.17 to 1.33

most_divergent <- variant_path_NES_padj %>% filter(NES_diff == max(NES_diff)) %>% dplyr::select(pathway, NES_diff)
most_convergent <- variant_path_NES_padj %>% filter(abs(NES_diff) == min(abs(NES_diff))) %>% dplyr::select(pathway, NES_diff)

most_divergent
most_convergent

classify_paths <- variant_path_NES_padj %>% mutate(
  converge_diverge =
    case_when(
      (NES_diff > 1.33 | NES_diff < 0.17) ~ "divergent",
      (NES_diff <= 1.33 & NES_diff >= 0.17) ~ "convergent"
    )
)
classify_paths

divergent_unfilt <- classify_paths %>% filter(converge_diverge == "divergent")
divergent_unfilt
nrow(divergent_unfilt)
convergent_unfilt <- classify_paths %>% filter(converge_diverge == "convergent")
convergent_unfilt
nrow(convergent_unfilt)
```


```{r visualize-NES-padj}

#determine padj filter by visualizing NES difference against -log10padj

rare_NES_padj <- ggplot(classify_paths, aes(x = NES_diff, y = -log10(rare_padj), color = converge_diverge)) +
  geom_jitter(alpha = 0.8) +
  scale_color_manual(values = c("convergent" = "lightblue", "divergent" = "#F4D03F")) +
  xlab("Difference in NES")+
  ylab("Rare Variant -log10(padj")+
  theme_classic()

common_NES_padj <- ggplot(classify_paths, aes(x = NES_diff, y = -log10(common_padj), color = converge_diverge)) +
  geom_jitter(alpha = 0.8) +
  scale_color_manual(values = c("convergent" = "lightblue", "divergent" = "#F4D03F")) +
  xlab("Difference in NES")+
  ylab("Common Variant -log10(padj")+
  theme_classic()

NES_padj <- rare_NES_padj + common_NES_padj

NES_padj
#ggsave("NES_distributions.png", plot = NES_hist, path = "../figures", width = 14, height = 10, dpi = 300)
```

## Apply FDR Pathway Filter

```{r apply-FDR-filter}
# apply a filter of padj < 0.25 to rare and common arms
rare_sig <- rare_pathways %>% filter(padj < 0.25) %>% pull(pathway)
common_sig <- common_pathways %>% filter(padj < 0.25) %>% pull(pathway)
both_sig <- intersect(rare_sig, common_sig)
all_paths_filt <- all_paths_overlap %>% filter(pathway %in% both_sig)
nrow(all_paths_filt)
setdiff(rare_sig, common_sig)  #rare only
setdiff(common_sig, rare_sig)  #common only
```


```{r pivot-wider-filt}
variant_path_NES <- all_paths_filt %>%
  dplyr::select(variant, pathway, NES) %>%
  pivot_wider(
    names_from = variant,
    values_from = NES,
    names_glue = "{variant}_NES"
  )

variant_path_NES
#sum(is.na(variant_path_NES))
```

## Calculate Filtered MAD

``` {r MAD-calc}
variant_path_NES <- variant_path_NES %>% mutate(
  NES_diff = rare_NES - common_NES, 
  MAD_filt = mad(NES_diff),
  median_NES_diff = median(NES_diff)) %>% relocate(NES_diff, .before = MAD_filt)

variant_path_NES

#unfiltered MA for robustness check: 0.5844998
mad_filt <- 0.2611755
median_filt <- 0.1677036
```

## Re-Classify

```{r classify-filt}
#threshold of 1 MAD away from the median for divergence:

#new median ± 1 MAD = 0.17 ± 0.26
#new convergence zone: -0.09 to 0.33

most_divergent <- variant_path_NES %>% filter(NES_diff == max(NES_diff)) %>% dplyr::select(pathway, NES_diff)
most_convergent <- variant_path_NES %>% filter(abs(NES_diff) == min(abs(NES_diff))) %>% dplyr::select(pathway, NES_diff)

most_divergent
most_convergent

classify_paths_filt <- variant_path_NES %>% mutate(
  converge_diverge =
    case_when(
      (NES_diff > .33 | NES_diff < -0.09) ~ "divergent",
      (NES_diff <= .33 & NES_diff >= -0.09) ~ "convergent"
    )
)
classify_paths_filt

divergent <- classify_paths_filt %>% filter(converge_diverge == "divergent")
divergent
nrow(divergent) # 8
convergent <- classify_paths_filt %>% filter(converge_diverge == "convergent")
convergent
nrow(convergent) # 7
```


```{r visualize-summary}
#inner join to add converge_diverge onto all_paths_overlap
classify_all_paths_overlap_filt <- inner_join(all_paths_filt, classify_paths_filt)
classify_all_paths_overlap_filt

p <- ggplot(classify_all_paths_overlap_filt, aes(x = NES, y = reorder(pathway, NES), fill = variant)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("common" = "lightblue", "rare" = "#F4D03F"))+
  facet_wrap(~ converge_diverge)+
  labs(
    title = "Convergence and Divergence in Enrichment Across Variant Type in Filtered Dataset",
    x = "Normalized Enrichment Score (NES)",
    y = NULL,
    fill = "Variant Type"
  ) +
  theme_minimal()

p

#ggsave("converge_diverge_enrichment.png", plot = p2, path = "../figures", width = 14, height = 10, dpi = 300)
```

```{r results}
#save results
path_results <- classify_paths_filt %>% dplyr::select(pathway, converge_diverge)
write.csv(path_results, "../results/path_results.csv")
```
## Session Info

```{r session-info}
sessionInfo()
```
