---
title: "SFARI Gene List Cross-Reference"
author: "Amalya Murrill"
date: "`r Sys.Date()`"
output: github_document
 
---
The SFARI database collects information about genes empirically implicated in ASD and categorizes them, specifically by how well-verified their mutations are in ASD risk. Our annotated genes were cross-referenced against this SFARI gene list, identifying 33% as SFARI-annotated.     


```{r knitting-rules}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Libraries

```{r libraries}
library(tidyverse)
```

## Import SFARI Genes

```{r sfari-import}
#import SFARI gene database data
sfari_data <- read.csv("../../data/raw/SFARI_genes.csv", header = TRUE)
colnames(sfari_data)
sfari_data <- sfari_data %>% dplyr::select(gene.symbol, gene.name, genetic.category, gene.score)
head(sfari_data)
```

## Import Annotated Pathways

```{r anno-import-and-select}
anno_results <- read.csv("../../results/loeuf_anno_results.csv", header = TRUE)
head(anno_results) 
```

## Join SFARI Data

```{r join-sfari}
#prepare annotated pathway data
anno_paths <- anno_results %>% 
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";")) %>%
  mutate(gene = sapply(leadingEdge, str_split ,";"))%>% 
  dplyr::select(-leadingEdge) %>% 
  unnest(gene) %>%
  relocate(c(gene, pathway), .before = 1)

nrow(anno_paths) # 1422 individual gene-pathway groupings

sfari_data_rename <- sfari_data %>%
  dplyr::rename(gene = gene.symbol)

cross_ref_inner <- anno_paths %>%
  dplyr::inner_join(sfari_data_rename, by = "gene")
cross_ref_inner
nrow(cross_ref_inner) # 470 individual gene-pathway groupings, 33% match with SFARI

cross_ref_outer <- anno_paths %>%
  dplyr::left_join(sfari_data_rename, by = "gene")
cross_ref_outer %>% filter(is.na(gene.name)) # perhaps genes of future interest?
```

## Explore Results

```{r top-explore}
#top 5 by padj
cross_ref_inner %>% arrange(padj) %>% distinct(pathway, .keep_all = TRUE) %>% head(5)
#top 5 by percent constrained
cross_ref_inner %>% arrange(desc(percent_constrained)) %>% distinct(pathway, .keep_all = TRUE) %>% head(5)
#how often genes appear (hub genes)
cross_ref_inner %>% count(gene, sort = TRUE)
```

## Session Info

```{r session-info}
sessionInfo()
```