---
title: "LOEUF Annotation and Gene-Level Constraint Assessment"
author: "Amalya Murrill"
date: "`r Sys.Date()`"
output: github_document
 
---
This script takes the enriched pathways from FGSEA that have been statistically categorized as convergent or divergent and extracts gene-level data. The findings are joined with corresponding LOEUF scores downloaded from gnomAD v2.1.1. Overall constraint for each pathway was assessed by calculating the proportion and then percentage of constrained genes (LOEUF < 0.35 as dictated by literature) within that pathway. The data was then formatted for Cytoscape output as a visualization of constraint within a network.   

```{r knitting-rules}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Libraries

```{r libraries}
library(tidyverse)
library(org.Hs.eg.db)
```

## Import FGSEA Results

```{r path-import}
#import leadingEdge data per pathway per variant
paths_data <- read.csv("../../results/all_path_data.csv", header = TRUE)

paths_data <- paths_data %>% dplyr::select(variant, pathway, padj,leadingEdge)
head(paths_data)
nrow(paths_data) # 38; 19 per variant

#import pathway classification 
classify_paths_results <- read.csv("../../results/path_results.csv", header = TRUE) 
classify_paths_results <- classify_paths_results %>% dplyr::select(pathway, converge_diverge)
nrow(classify_paths_results) # 19
head(classify_paths_results)
```

## Join Gene Data with Convergence Data

```{r classify-genes}
classify_genes <- left_join(paths_data, classify_paths_results, by = "pathway")
head(classify_genes)
``` 

### Convert EntrezIDs to Gene Symbols

```{r rename-genes}
leadingEdgeGenes <- classify_genes %>% dplyr::select(leadingEdge) %>% pull()

#split each list into individual gene characters
leadingEdgeGenes_split <- strsplit(leadingEdgeGenes, ";")

#use lapply across mapping 
symbols <- lapply(leadingEdgeGenes_split, function(r) mapIds(
  org.Hs.eg.db,
  keys = r,
  column = "SYMBOL",
  keytype = "ENTREZID",
  multiVals = "first"
))

#rename
symbols <- lapply(symbols, unname)
head(symbols)

#re-add to table
classify_genes <- classify_genes %>% mutate(leadingEdge = symbols)
head(classify_genes)
``` 

## Import gnomAD Constraint Scores

```{r gnomAD-import}
gnomad <- read.csv("../../data/raw/gnomad.v2.1.1.lof_metrics.by_gene.txt.bgz", header = TRUE, sep = "\t")
head(gnomad)
colnames(gnomad) #we want "gene" and "oe_lof_upper"

gnomad_loeuf <- gnomad %>% dplyr::select(gene, oe_lof_upper)
head(gnomad_loeuf)
```

## Annotate Genes with LOEUF Scores

```{r annotate}
#make named vector
loeuf <- deframe(gnomad_loeuf[ ,c("gene", "oe_lof_upper")])
head(loeuf)

#match to symbols
loeuf_scores <- lapply(symbols, function(x)loeuf[x])
length(loeuf_scores) == length(symbols) #matched
loeuf_scores

#re-add to table
loeuf_scores <- lapply(loeuf_scores, unname)
classify_genes <- classify_genes %>% mutate(LOEUF = loeuf_scores)
head(classify_genes)
```

## Assess Constraint

```{r calculate-contraint}
#determine constraint by calculating proportion of genes with LOEUF scores below 0.35 threshold
#calculate proportion ignoring NA values
classify_genes <- classify_genes %>% mutate(percent_constrained = sapply(LOEUF, function(x)((sum(x < 0.35, na.rm = TRUE)/sum(!is.na(x)))*100))) %>% # remove NAs
relocate(percent_constrained, .before = LOEUF)
classify_genes
```

```{r explore-constraint-results}
#explore results
classify_genes %>% arrange(desc(percent_constrained)) # arrange by constraint desc
classify_genes %>% arrange(desc(percent_constrained)) %>% filter(converge_diverge == "convergent")
classify_genes %>% arrange(desc(percent_constrained)) %>% filter(converge_diverge == "divergent")

collapsed_classified_genes <- classify_genes %>%
  dplyr::select(-LOEUF) %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))

#write.csv(collapsed_classified_genes, "../../results/loeuf_anno_results.csv", row.names = FALSE)
```

## Cytoscape File Output

```{r cytoscape-files}
#select top constrained pathways
top <- classify_genes %>% arrange(desc(percent_constrained)) %>% distinct(pathway, .keep_all = TRUE) %>% head(10)

#pathway level metadata
top_node_attributes <- top %>% 
  dplyr::select(pathway, variant, converge_diverge, percent_constrained) %>%
  distinct() %>% 
  mutate(name = paste0(pathway, "_", variant)) #unique ID for pathway

#edge list (pathway-variant → gene)
top_edge_list <- top %>% 
  dplyr::select(pathway, variant, leadingEdge) %>%
  unnest(leadingEdge) %>% 
  mutate(
    source = paste0(pathway, "_", variant),
    target = leadingEdge) %>%
  dplyr::select(source, target)

#pathway nodes
pathway_nodes <- top_edge_list %>%
  distinct(source) %>%
  dplyr::rename(name = source) %>%
  mutate(node_type = "pathway") %>%
  left_join(top_node_attributes, by = "name")

#gene nodes
gene_nodes <- top_edge_list %>%
  distinct(target) %>%
  dplyr::rename(name = target) %>%
  mutate(node_type = "gene")

#combine nodes
top_all_nodes <- dplyr::bind_rows(pathway_nodes, gene_nodes)

#write CSVs
#write.csv(top_all_nodes, "../../results/cyto_top_all_nodes.csv", row.names = FALSE, quote = FALSE)
#write.csv(top_edge_list, "../../results/cyto_top_edge_list.csv", row.names = FALSE, quote = FALSE)
```

## Session Info

```{r session-info}
sessionInfo()
```